---
title: "Integrating heatwave time series respirometry data"
author: "Ariana S Huffmyer, E5 RoL Team"
edited by: "DM Becker-Polinski"
date: "20220830"
output: html_document
editor_options: 
  chunk_output_type: console
--- 
NOTE: If you want to run this script, do not Knit, instead "run all chunks below" and look at console output.

## Script to run all physiological and biological data used for the heatwave timeseries. https://github.com/daniellembecker/Gametogenesis/tree/main/heatwave/timeseries
 
# Set Up    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
## install packages if you dont already have them
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("RColorBrewer")) install.packages("RColorBrewer")
if (!require("lme4")) install.packages("lme4")
if (!require("lmerTest")) install.packages("lmerTest")
if (!require("car")) install.packages("car")
if (!require("effects")) install.packages("effects")
if (!require("ggfortify")) install.packages("ggfortify")
if (!require("cowplot")) install.packages("cowplot")
if (!require("vegan")) install.packages("vegan")
if (!require("corrr")) install.packages("corrr")
if (!require("ggcorrplot")) install.packages("ggcorrplot")
if (!require("GGally")) install.packages("GGally")
if (!require("broom")) install.packages("broom")
if (!require("cowplot")) install.packages("cowplot")
if (!require("emmeans")) install.packages("emmeans")

# load packages
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(ggfortify)
library(cowplot)
library(vegan)
library(corrr)
library(ggcorrplot)
library(multcompView)
library(GGally)
library(broom)
library(patchwork)
library(cowplot)
library(emmeans)

```
 
# Load and manipulate data for photosynthesis and respiration measurements from heatwave

## Loading data files for PR rates   

Load all .csv files from output of all timepoints for each biological response     
 
```{r}
PR_files <- list.files("../", pattern = "PR_rates.csv", recursive=T, full.names=T)

```
 
## Read data files 
 
Load all data frames.  

```{r}
#PR rates
PR_dataset <- data.frame() #make dataframe
PR_dataset <- PR_dataset[-c(1)] #remove first X column

for (i in 1:length(PR_files)){
  PR_df <- read.csv(PR_files[i]) #each file will be read in
  PR_dataset <- rbind(PR_dataset, PR_df) #for each iteration, bind the new data to the building dataset
}

#provide context for Rd and P responses by light levels in data 
PR_dataset <- PR_dataset %>% 
  mutate(Assay = case_when(
    Light_Value == "0" ~ "Rd",
    Light_Value == "655" ~ "P",
    Light_Value == "657" ~ "P",
    Light_Value == "610" ~ "P",
    Light_Value == "625" ~ "P",
    Light_Value == "615" ~ "P",
    Light_Value == "656" ~ "P",
    Light_Value == "643" ~ "P",
    Light_Value == "651" ~ "P",
    Light_Value == "641" ~ "P",
    Light_Value == "649" ~ "P", 
    Light_Value == "654" ~ "P",
    Light_Value == "655" ~ "P",
    Light_Value == "621" ~ "P",
    Light_Value == "614" ~ "P",
    Light_Value == "630" ~ "P",
    Light_Value == "654" ~ "P", 
    Light_Value == "645" ~ "P",
    Light_Value == "650" ~ "P",
    Light_Value == "657" ~ "P",
    Light_Value == "658" ~ "P",
    Light_Value == "631" ~ "P",
    Light_Value == "649" ~ "P",
    Light_Value == "655" ~ "P"))

```

```{r}
#remove files that are not needed from loops
rm(list = ls(pattern = "*_df"))
```

## Generate master data frame    

Read in tag metadata frame. 
```{r}
#Load tag metadata sheet 
tags <- read.csv("fragment_ metadata.csv")
```


# Organize data frame for respiration and photosynthesis univariate analysis

```{r}
# change from long to wide data format
data.org <- PR_dataset %>%
  pivot_wider(names_from = Assay, values_from = micromol.cm2.h)

#remove first X column
data.org <- data.org[-c(1)] 

#remove date column to remove NAs later
data.org <- data.org[, -6]  

#Summarize all response values by group (fragment ID + timepoint + Treatment).
summary_fragments <- group_by(data.org, fragment_ID, timepoint, Treatment) %>%
  summarise(across(.cols=everything(), ~mean(.x, na.rm = TRUE)))%>%
  drop_na("timepoint")%>%
  write_csv(., "timeseries_analysis/Output/summary_respo_fragment_ID.csv")

# Add in tank.num metadata 
summary_fragments <- left_join(summary_fragments, tags)

#remove notes column
summary_fragments <- summary_fragments[, -17]  

#make respiration positive 
summary_fragments$Rd <- -summary_fragments$Rd

#make new column for gross photosynthesis and add net photosynthesis (the total rate of carbon fixation (reduction of CO2) without considering that some of the CO2 is lost in respiration) + the absolute value of Rd for gross photosynthesis - the total rate of carbon fixation (reduction of CO2) without considering that some of the CO2 is lost in respiration
summary_fragments$GP <- summary_fragments$P + summary_fragments$Rd

#Calculate a P:R ratio = gross photosynthesis / respiration and make new column. A 1:1 ratio indicates that photosynthesis meets respiratory demand.    
summary_fragments$GP.R <- abs(summary_fragments$GP)/abs(summary_fragments$Rd)
#calculate ratio with absolute values
#remove outliers detected by values of P:R ratio data


# Mapping of timepoints to corresponding dates
summary_fragments <- summary_fragments %>%
  mutate(formatted_date = case_when(
    timepoint == "TP1" ~ "03/19",
    timepoint == "TP2" ~ "03/26",
    timepoint == "TP3" ~ "03/31",
    timepoint == "TP4" ~ "04/10",
    timepoint == "TP5" ~ "04/18",
    TRUE ~ as.character(timepoint)  # Keep original date for other cases if needed
  ))

#summarize data for each timepoint
complete_sum <-summary_fragments%>%
  group_by(Treatment, timepoint)%>%
  summarise(across(.cols=everything(), ~mean(.x, na.rm = TRUE)))%>%
  drop_na("timepoint")


#remove columns with no data
complete_sum <- complete_sum[, -c(3, 10:13, 15)] %>%
  write_csv(., "timeseries_analysis/Output/summary_respo_timepoint.csv")

```

# Remove initial timepoint from analysis because we want to see differences in treatments across timepoints

```{r}
summary_fragments.filtered <- summary_fragments %>%
  filter(!is.na(treatment) & treatment != "initial") 


```


### Analysis of NP   

Build a mixed model for univariate analysis and examine data distribution. 

`P_model<-lmer(P~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )` 

```{r}
P_model<-lmer(P~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(P_model))
```

Residuals are not normally distributed. Attempt with log transformation.      

`P_model<-lmer(log(P~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered ))` 

```{r}
P_model<-lmer(log(P)~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(P_model))
```

Generate a Type III Anova of model.    

```{r}
anova(P_model, type="III")
summary(P_model)

# run post hoc on timepoints and treatment
emm_P <-emmeans(P_model, ~ timepoint)
pairs(emm_P)

# Generate letter-based groupings
cld_result_P <- multcomp::cld(emm_P, Letters = letters)
letters_data_P_og <- as.data.frame(cld_result_P)

# Calculate means per treatment to calculate percent change
means_per_treatment <- summary_fragments.filtered  %>%
  group_by(treatment) %>%
  summarise(mean_p = mean(P))

means_per_treatment
```

Timepoint x treatment is significant and main effect of timepoint

Type III Analysis of Variance Table with Satterthwaite's method
                    Sum Sq Mean Sq NumDF DenDF F value   Pr(>F)   
timepoint           1.1559 0.38531     3    66  1.7770 0.160106   
Treatment           0.1024 0.10240     1    22  0.4722 0.499139   
timepoint:Treatment 2.7971 0.93237     3    66  4.3000 0.007842 **

### Analysis of Rd   

Build a mixed model for univariate analysis and examine data distribution. 

`Rd_model<-lmer(Rd~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )` 

```{r}
rd_model<-lmer(Rd~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(rd_model))
```

Residuals are not normally distributed. Attempt with log transformation.      

`rd_model<-lmer(log(Rd~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered ))` 

```{r}
rd_model<-lmer(log(Rd)~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(rd_model))
```

Generate a Type III Anova of model.    

```{r}
anova(rd_model, type="III")

# run post hoc on timepoints and treatment
emm_rd <-emmeans(rd_model, ~ timepoint)
pairs(emm_rd)

# Generate letter-based groupings
cld_result_rd <- multcomp::cld(emm_rd, Letters = letters)
letters_data_rd_og <- as.data.frame(cld_result_rd)

# Calculate means per treatment to calculate percent change
means_per_treatment <- summary_fragments.filtered %>%
  group_by(treatment) %>%
  summarise(mean_R = mean(Rd))

means_per_treatment

```

Interaction is not significant but main effects of timepoint and treatment are significant:

Type III Analysis of Variance Table with Satterthwaite's method
                     Sum Sq Mean Sq NumDF DenDF F value   Pr(>F)   
timepoint           1.27540 0.42513     3    87  4.5218 0.005388 **
Treatment           0.47645 0.47645     1    87  5.0676 0.026897 * 
timepoint:Treatment 0.69520 0.23173     3    87  2.4648 0.067653 .


### Analysis of GP

Build a mixed model for univariate analysis and examine data distribution. 

`GP_model<-lmer(GP~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )` 

```{r}
GP_model<-lmer(GP~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(GP_model))
```

Residuals are not normally distributed. Attempt with log transformation.      

`rd_model<-lmer(log(Rd~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered ))` 

```{r}
GP_model<-lmer(log(GP)~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(GP_model))
```


Generate a Type III Anova of model.    

```{r}
anova(GP_model, type="III")
summary(GP_model)

# run post hoc on timepoints and treatment
emm_GP <-emmeans(GP_model, ~ timepoint)
pairs(emm_GP)

# Generate letter-based groupings
cld_result_GP <- multcomp::cld(emm_GP, Letters = letters)
letters_data_GP_og <- as.data.frame(cld_result_GP)

# Calculate means per treatment to calculate percent change
means_per_treatment <- summary_fragments.filtered %>%
  group_by(treatment) %>%
  summarise(mean_GP = mean(GP))

means_per_treatment
```

Interaction is significant 

Type III Analysis of Variance Table with Satterthwaite's method
                     Sum Sq Mean Sq NumDF DenDF F value   Pr(>F)   
timepoint           0.38293 0.12764     3    66  1.3751 0.258080   
Treatment           0.18574 0.18574     1    22  2.0010 0.171197   
timepoint:Treatment 1.25354 0.41785     3    66  4.5015 0.006195 **

### Analysis of GP:R  

Build a mixed model for univariate analysis and examine data distribution. 

`GP.R_model<-lmer(GP.R~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )` 

```{r}
GP.R_model<-lmer(GP.R~timepoint*Treatment+(1|colony_ID), na.action=na.omit, data=summary_fragments.filtered )
qqPlot(residuals(GP.R_model))
```

#residuals are normal move ahead

Generate a Type III Anova of model.    

```{r}
anova(GP.R_model, type="III")

# run post hoc on timepoints
emm_GPR<-emmeans(GP.R_model, ~ timepoint)
pairs(emm_GPR)

# Generate letter-based groupings
cld_result_GPR <- multcomp::cld(emm_GPR, Letters = letters)
letters_data_GPR_og <- as.data.frame(cld_result_GPR)

```

Interaction is not significant but main effects of timepoints are significant:

Type III Analysis of Variance Table with Satterthwaite's method
                     Sum Sq Mean Sq NumDF DenDF F value    Pr(>F)    
timepoint           1.79164 0.59721     3    66  8.5731 6.915e-05 ***
Treatment           0.26241 0.26241     1    22  3.7670   0.06519 .  
timepoint:Treatment 0.17285 0.05762     3    66  0.8271   0.48364 



Make net photosynthesis plot
```{r}
# Extract relevant columns from the compact letter display for post hoc data
letters_data_NP <- letters_data_P_og %>%
  select(timepoint, .group)

# Create a new row for TP1
new_row <- data.frame(
  timepoint = "TP1",
  .group = "na"
)

# Append the new row to the dataframe
letters_data_NP <- rbind(letters_data_NP, new_row)

# Prepare your plotting data by filtering and merging with the letters_data_sym
plot_data_NP <- summary_fragments %>%
  filter(!is.na(P)) %>%
  filter(!is.na(timepoint)) %>%
  select(fragment_ID, timepoint, Treatment, P, formatted_date) %>%
  left_join(letters_data_NP, by = "timepoint")

# check for and remove NP outliers
NP_clean <- plot_data_NP %>%
  filter(P < 0.80) %>% filter(P > 0) 

# Summarize the CLD data to get one letter per timepoint
cld_summary_NP <- NP_clean %>%
  group_by(timepoint, formatted_date, Treatment) %>%
  summarise(.group = first(.group), max_y = max(P)) %>%
  distinct(timepoint, .keep_all = TRUE)  # Keep only one row per timepoint

# Remove any additional attributes (if any) while keeping it a dataframe
cld_summary_NP <- as.data.frame(cld_summary_NP)

#make NP plot
NP_plot <- ggplot(NP_clean, aes(x = formatted_date, y = P, fill = Treatment)) +
    geom_boxplot(outlier.size = 0) +
    geom_point(pch = 21, size=2, position = position_jitterdodge(0.2)) + 
    xlab("Timepoint") + 
    geom_text(data = cld_summary_NP, aes(x = formatted_date, y = max_y + 0.05 * max_y, label = .group), size = 8, vjust = -0.5,fontface = "bold") + # Add letters
    ylim(0,1) +
ylab(expression(bold(paste("Net P (" *mu*"mol ", O[2] ," " *cm^-2 *hr^-1*")"))))+
scale_fill_manual(values = c("darkblue", "red", "gray")) + #change gradient color 
  theme_classic() + 
    theme(
      legend.position = c(0.9, 0.85),
      legend.title=element_text(face="bold", size=40),
      legend.text=element_text(size=40),
      axis.title=element_text(face="bold", size=40),
      axis.text=element_text(size=40, color="black"), 
      strip.text.x=element_text(face="italic", size=40)
      ); NP_plot

```

Make respiration plot
```{r}
# Extract relevant columns from the compact letter display for post hoc data
letters_data_RD <- letters_data_rd_og %>%
  select(timepoint, .group)

# Create a new row for TP1
new_row <- data.frame(
  timepoint = "TP1",
  .group = "na"
)

# Append the new row to the dataframe
letters_data_RD <- rbind(letters_data_RD, new_row)

# Prepare your plotting data by filtering and merging with the letters_data_sym
plot_data_RD <- summary_fragments %>%
  filter(!is.na(Rd)) %>%
  filter(!is.na(timepoint)) %>%
  select(fragment_ID, timepoint, Treatment, Rd, formatted_date) %>%
  left_join(letters_data_RD, by = "timepoint")

#check for and remove Rd outliers
Rd_clean <- plot_data_RD %>%
  filter(Rd < 0.97) %>%
  filter(Rd > 0.0)

# Summarize the CLD data to get one letter per timepoint
cld_summary_RD <- Rd_clean %>%
  group_by(timepoint, formatted_date, Treatment) %>%
  summarise(.group = first(.group), max_y = max(Rd)) %>%
  distinct(timepoint, .keep_all = TRUE)  # Keep only one row per timepoint

# Remove any additional attributes (if any) while keeping it a dataframe
cld_summary_RD <- as.data.frame(cld_summary_RD)

#make Rd plot
Rd_plot <- ggplot(Rd_clean, aes(x = formatted_date, y = Rd, fill = Treatment)) +
    geom_boxplot(outlier.size = 0) +
    geom_point(pch = 21, size=2, position = position_jitterdodge(0.2)) + 
    xlab("Timepoint") + 
    geom_text(data = cld_summary_RD, aes(x = formatted_date, y = max_y + 0.05 * max_y, label = .group), size = 8, vjust = -0.5,fontface = "bold") + # Add letters
    ylab(expression(bold(paste("Respiration Rate (" *mu*"mol ", O[2] ," " *cm^-2 *h^-1*")"))))+
   scale_fill_manual(values = c("darkblue", "red", "gray")) + #change gradient color 
  theme_classic() + 
    theme(
      legend.position="none",
      legend.title=element_text(face="bold", size=40),
      legend.text=element_text(size=40),
      axis.title=element_text(face="bold", size=40),
      axis.text=element_text(size=40, color="black"), 
      strip.text.x=element_text(face="italic", size=40)
      ); Rd_plot

```


Make gross photosynthesis plot
```{r}
# Extract relevant columns from the compact letter display for post hoc data
letters_data_GP <- letters_data_GP_og %>%
  select(timepoint, .group)

# Create a new row for TP1
new_row <- data.frame(
  timepoint = "TP1",
  .group = "na"
)

# Append the new row to the dataframe
letters_data_GP <- rbind(letters_data_GP, new_row)

# Prepare your plotting data by filtering and merging with the letters_data_sym
plot_data_GP <- summary_fragments %>%
  filter(!is.na(GP)) %>%
  filter(!is.na(timepoint)) %>%
  select(fragment_ID, timepoint, Treatment, GP, formatted_date) %>%
  left_join(letters_data_GP, by = "timepoint")

#check for and remove GP outliers
GP_clean <- plot_data_GP %>%
  filter(GP < 2.0) %>%
  filter(GP > 0.3)

# Summarize the CLD data to get one letter per timepoint
cld_summary_GP <- GP_clean  %>%
  group_by(timepoint, formatted_date, Treatment) %>%
  summarise(.group = first(.group), max_y = max(GP)) %>%
  distinct(timepoint, .keep_all = TRUE)  # Keep only one row per timepoint

# Remove any additional attributes (if any) while keeping it a dataframe
cld_summary_GP <- as.data.frame(cld_summary_GP)

#make GP plot
GP_plot <- ggplot(GP_clean, aes(x = formatted_date, y = GP, fill = Treatment)) +
    geom_boxplot(outlier.size = 0) +
    geom_point(pch = 21, size=2, position = position_jitterdodge(0.2)) + 
    xlab("Timepoint") + 
    geom_text(data = cld_summary_GP, aes(x = formatted_date, y = max_y + 0.05 * max_y, label = .group), size = 8, vjust = -0.5,fontface = "bold") + # Add letters
    ylab(expression(bold(paste("Gross Photosynthetic Rate (" *mu*"mol ", O[2] ," " *cm^-2 *h^-1*")"))))+
   scale_fill_manual(values = c("darkblue", "red", "gray")) + #change gradient color 
  theme_classic() + 
    theme(
      legend.position="none",
      legend.title=element_text(face="bold", size=40),
      legend.text=element_text(size=40),
      axis.title=element_text(face="bold", size=40),
      axis.text=element_text(size=40, color="black"), 
      strip.text.x=element_text(face="italic", size=40)
      ); GP_plot
```


Make GP:R plot
```{r}
# Extract relevant columns from the compact letter display for post hoc data
letters_data_GP_R <- letters_data_GPR_og %>%
  select(timepoint, .group)

# Create a new row for TP1
new_row <- data.frame(
  timepoint = "TP1",
  .group = "na"
)

# Append the new row to the dataframe
letters_data_GPR <- rbind(letters_data_GP_R , new_row)

# Prepare your plotting data by filtering and merging with the letters_data_sym
plot_data_GPR <- summary_fragments %>%
  filter(!is.na(GP.R)) %>%
  filter(!is.na(timepoint)) %>%
  select(fragment_ID, timepoint, Treatment, GP.R, formatted_date) %>%
  left_join(letters_data_GPR, by = "timepoint")

# Summarize the CLD data to get one letter per timepoint
cld_summary_GPR <- plot_data_GPR  %>%
  group_by(timepoint, formatted_date, Treatment) %>%
  summarise(.group = first(.group), max_y = max(GP.R)) %>%
  distinct(timepoint, .keep_all = TRUE)  # Keep only one row per timepoint

# Remove any additional attributes (if any) while keeping it a dataframe
cld_summary_GPR <- as.data.frame(cld_summary_GPR)

#make GP:R plot
GP_R_plot <- ggplot(plot_data_GPR, aes(x = formatted_date, y = GP.R, fill = Treatment)) +
    geom_boxplot(outlier.size = 0) +
    geom_point(pch = 21, size=2, position = position_jitterdodge(0.2)) + 
    xlab("Timepoint") + 
    geom_text(data = cld_summary_GPR, aes(x = formatted_date, y = max_y + 0.05 * max_y, label = .group), size = 8, vjust = -0.5,fontface = "bold") + # Add letters
    ylab(expression(bold(paste("GP:R (" *mu*"mol ", O[2] ," " *cm^-2 *h^-1*")"))))+
   scale_fill_manual(values = c("darkblue", "red", "gray")) + #change gradient color 
  theme_classic() + 
    theme(
      legend.position="none",
      legend.title=element_text(face="bold", size=40),
      legend.text=element_text(size=40),
      axis.title=element_text(face="bold", size=40),
      axis.text=element_text(size=40, color="black"), 
      strip.text.x=element_text(face="italic", size=40)
      ); GP_R_plot

```


Join all plots together.  
```{r}

all_resp_Figure<-plot_grid(NP_plot, Rd_plot, GP_plot, GP_R_plot, ncol=2, nrow=2, labels = c('A', 'B', 'C', 'D'),rel_heights= c(1,1,1,1), rel_widths = c(1,1,1,1), label_y=1, align="h", label_size = 50)

ggsave(filename="timeseries_analysis/Figures/all_resp_figure.pdf", plot=all_resp_Figure, dpi=500, width=30, height=30, units="in")
ggsave(filename="timeseries_analysis/Figures/all_resp_figure.png", plot=all_resp_Figure, dpi=500, width=30, height=30, units="in")
```

```{r}

NP_Rd <-plot_grid(NP_plot, Rd_plot, ncol=2, nrow=1, labels = c('A', 'B'), label_size = 60)

ggsave(filename="timeseries_analysis/Figures/NP_Rd_Figure.png", plot=NP_Rd, dpi=500, width=35, height=20, units="in")
ggsave(filename="timeseries_analysis/Figures/NP_Rd_Figure.pdf", plot=NP_Rd, dpi=500, width=35, height=20, units="in")
```


