---
title: "Reproductive Ecology Analysis: spawning success, egg size, eggs per bundle, and fertilization success"
author: "DM Becker-Polinski"
date: "2023" 
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

This script analyzes spawning success, egg size, and fertilization success as physiology metrics in the *A. pulchra* heatwave time series and reproductive ecology between control and heatwave colonies during spawning October 2022.    

# **Setup**  

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r setup, include = FALSE}
## install packages if you dont already have them
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("RColorBrewer")) install.packages("RColorBrewer")
if (!require("car")) install.packages("car")
if (!require("emmeans")) install.packages("emmeans")
if (!require("multcomp")) install.packages("multcomp")
if (!require("knitr")) install.packages("knitr")
if (!require("kableExtra")) install.packages("kableExtra")
if (!require("lme4")) install.packages("lme4")
if (!require("GGally")) install.packages("GGally")
if (!require("lmerTest")) install.packages("lmerTest")
if (!require("cowplot")) install.packages("cowplot")
if (!require("PNWColors")) install.packages("PNWColors")

# load packages
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(car)
library(emmeans)
library(multcomp)
library(knitr)
library(kableExtra)
library(lme4)
library(GGally)
library(lmerTest)
library(cowplot)
library(patchwork)
library(ggsignif)
library(Hmisc)
library(PNWColors)
```

In this project, physiological assays include egg size, fertilization success.  

# **1. Egg Size/Volume**  

First, plot and analyze egg size measured as volume. Volume is calculated as a prolate spheroid using the length and width measurements of each individual sampled across heatwave spawning project.     

## Import and manipulate data

Load in data.
```{r}
# Egg size data
size <- read_csv("data/reproductive_ecology/egg_volume.csv")

# select columns
size <- size %>%
  dplyr::select(tube.ID, treatment, replicate, type, `width (mm)`, `length (mm)`)%>%
  rename(width=`width (mm)`, length = `length (mm)`)%>%
  drop_na() #remove na's that could not be measured

size$replicate<-as.factor(size$replicate)
size$tube.ID<-as.factor(size$tube.ID)
size$treatment<-as.factor(size$treatment)
size$type<-as.factor(size$type)

```

Remove na's from data frame.  
```{r}
# Calculate mean counts for each sample
size <- size %>%
  dplyr::select(tube.ID, treatment, replicate, type, width, length)%>%
  drop_na()
```


## Calculate volume  

Calculate the volume of each treatment. For eggs, we will use a spheroid equation. This best approximates the shape of each life stage.  

Spheroid: V = 4/3 * pi * a^3 where a is 1/2 the width (radius)

Calculate the volume for each individual as a spheroid. Using this equation: V = 4/3 * pi * a^3 where a is 1/2 the width (radius)
```{r}
# Define a function to calculate the volume based on life stage
calculate_volume <- function(type, length, width) {
  case_when(
    type %in% c("egg") ~ (4/3) * pi * (width/2)^3,
    TRUE ~ NA_real_
  )
}

# Use mutate function to calculate volume based on life stage
size <- size %>% mutate(volume = calculate_volume(type, length, width))

# Print the updated dataframe
print(size)
```

## Plot larval volume    

Plot data with mean and standard error for each treatment 
```{r}
size %>%
  ggplot(aes(x = treatment, y = volume, color = type)) +
  labs(x = "",y = "Mean Larval Volume (mm^3)") +
  geom_jitter(width = 0.1) +                                            # Plot all points
  stat_summary(fun.data = mean_cl_normal, fun.args = list(mult = 1),    # Plot standard error
               geom = "errorbar", color = "black", width = 0.5) +
  stat_summary(fun = mean, geom = "point", color = "black") + # Plot mean
  theme_classic()
```

Remove the outlier in the metamorphosed polyp group. 
```{r}
size<-size%>%
  filter(volume<0.3)
```

Present means and standard error of each group and save summary table.  

```{r}
size%>%
  group_by(treatment)%>%
  summarise(n=length(volume),
            Mean=format(round(mean(volume), 3), 3), 
            SE=format(round(sd(volume)/sqrt(length(volume)),3),3))%>%
  dplyr::rename(Treatment=treatment)%>%
  kbl(caption="Descriptive statistics of larval volume across ontogeny")%>%
  kable_classic(full_width=FALSE, html_font="Arial")%>%
  row_spec(0, bold = TRUE) 

#need to output to csv 
size%>%
  group_by(treatment)%>%
  summarise(n=length(volume),
            Mean=format(round(mean(volume), 3), 3), 
            SE=format(round(sd(volume)/sqrt(length(volume)),3),3))%>%
  dplyr::rename(Treatment=treatment)%>%
  write_csv(., "output/reproductive_ecology/egg_volume_table.csv")
```

Plot data as a box plot.     
```{r}
size_plot<-size %>%
    ggplot(., aes(x = treatment, y = volume, colour=type)) +
    geom_boxplot(aes(colour=treatment), lwd=1) +
    geom_point(aes(fill=treatment), colour="black", pch = 21, size=4, position = position_jitterdodge(0.5), alpha=0.6) + 
    xlab("") + 
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Egg Size (mm"^3, ")"))))+
  xlab(expression(bold(paste("Treatment"))))+
    scale_x_discrete(labels=c("ambient" = "Ambient", "hot" = "Hot")) +
    ggtitle("Kruskal-Wallis, p = 0.0001") +
      annotate("text", x = 1.45, y = 0.220, label = "*", size = 12) +
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=16, color="black"),
      axis.text=element_text(size=14, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=14)); size_plot

ggsave(plot=size_plot, "figures/reproductive_ecology/egg_size.png",height=8, width=8)
ggsave(plot=size_plot, "figures/reproductive_ecology/egg_size.pdf",height=8, width=8)
```


Plot data as a mean and standard error plot.     
```{r}

size_plot_mean<-size %>%
  group_by(treatment)%>%
  summarise(mean=mean(volume, na.rm=TRUE), 
            n=length(volume),
            se=sd(volume, na.rm=TRUE)/sqrt(length(volume)))%>%

    ggplot(., aes(x = treatment, y = mean)) +
    geom_point(aes(fill=treatment), colour="black", pch = 21, size=8, alpha=1) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se, group=treatment), width=0, linetype="solid", position=position_dodge(0.4), size=0.8, color="black")+
    xlab("") + 
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Volume (mm"^3, ")"))))+
   xlab(expression(bold(paste("Treatment"))))+
    theme_classic() + 
    annotate("text", x = 1.45, y = 0.110, label = "*", size = 12) +
    ylim(0,0.12)+
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=16, color="black"),
      axis.text=element_text(size=14, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=14), 
      axis.text.x=element_text(angle=45, hjust=1, color="black")
      ); size_plot_mean

ggsave(plot=size_plot_mean, "figures/reproductive_ecology/egg_size_mean.png",height=6, width=6)
ggsave(plot=size_plot_mean, "figures/reproductive_ecology/egg_size_mean.pdf",height=6, width=6)
```

## Statistical analysis  

Run ANOVA test on egg size. 
```{r}
size_model<-aov(volume~treatment, data=size)
summary(size_model)
qqPlot(residuals(size_model))
leveneTest(residuals(size_model)~treatment, data=size)
```

There appears to be a violation in variance assumptions but not normality. Conduct non-parametric test (Kruskal Wallis).  

```{r}
kruskal.test(volume~treatment, data=size)
``` 

There is a significant difference in egg size between groups with higher from hot treatment.

Kruskal-Wallis rank sum test

data:  volume by treatment
Kruskal-Wallis chi-squared = 41.682, df = 1, p-value = 1.074e-10






# **2. Fertilization Success**  

First, plot and analyze 2, 4, and more cell fertilizartion success propotions. The fertilization success rate is determined for each concentration as the quantity of normally fertilized eggs, divided by the total eggs added.  

## Import and manipulate data

Load in data.
```{r}
# fertilization success data
fert <- read_csv("data/reproductive_ecology/fertilization_success.csv")

#make all columns factors
# fert$tube.ID<-as.factor(fert$tube.ID)
# fert$num.eggs<-as.factor(fert$num.eggs)
# fert$egg<-as.factor(fert$egg)
# fert$two.cells<-as.factor(fert$two.cells)
# fert$four.cells<-as.factor(fert$four.cells)
# fert$more.cells<-as.factor(fert$more.cells)
# fert$deformed.embryos<-as.factor(fert$deformed.embryos)

```


## Calculate fertilization success rate (%) for each morph 

Calculate the fertilization success rate for each stage of cleavage. 
```{r}
# Define a function to calculate the fert success at each stage or section
#make fert a data frame due to error in ops.data.frame error from calculation in line 267
fert <- as.data.frame(fert)

#calcualte fert success per clev stage, using mutate to take columns from egg to deformed embryos, and divide each column by the total number of eggs column * 100 to get fert success per clevage stage
calculate_fert_success <- fert %>% mutate_at(vars(egg:deformed.embryos),list(prop=~./num.eggs*100))

# Print the updated dataframe
print(calculate_fert_success)
```

## Plot total fertilization success by treatment  

Plot data with mean and standard error for each treatment 
```{r}
calculate_fert_success %>%
  ggplot(aes(x = treatment, y = tot.fert.success)) +
  labs(x = "",y = "Fertilization Success Rate (%)") +
  geom_jitter(width = 0.1) +                                            # Plot all points
  stat_summary(fun.data = mean_cl_normal, fun.args = list(mult = 1),    # Plot standard error
               geom = "errorbar", color = "black", width = 0.5) +
  stat_summary(fun = mean, geom = "point", color = "black") + # Plot mean
  theme_classic()
```

Present means and standard error of each group and save summary table.  

```{r}
calculate_fert_success%>%
  group_by(treatment)%>%
  summarise(n=length(tot.fert.success),
            Mean=format(round(mean(tot.fert.success), 3), 3), 
            SE=format(round(sd(tot.fert.success)/sqrt(length(tot.fert.success)),3),3))%>%
  dplyr::rename(Treatment=treatment)%>%
  kbl(caption="Descriptive statistics of fertilization success rate between treatments")%>%
  kable_classic(full_width=FALSE, html_font="Arial")%>%
  row_spec(0, bold = TRUE) 

#need to output to csv 
calculate_fert_success%>%
  group_by(treatment)%>%
  summarise(n=length(tot.fert.success),
            Mean=format(round(mean(tot.fert.success), 3), 3), 
            SE=format(round(sd(tot.fert.success)/sqrt(length(tot.fert.success)),3),3))%>%
  dplyr::rename(Treatment=treatment)%>%
  write_csv(., "output/reproductive_ecology/total.fert.success.csv")
```

Plot data as a box plot.     
```{r}
total_fert_success_plot <- calculate_fert_success %>%
    ggplot(., aes(x = treatment, y = tot.fert.success, color = treatment)) +
    geom_boxplot(aes(colour=treatment), lwd=1) +
    geom_point(aes(fill=treatment), colour="black", pch = 21, size=4, position = position_jitterdodge(0.5), alpha=0.6) + 
    xlab("") + 
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Fertilization Success Rate (%)"))))+
  xlab(expression(bold(paste("Treatment"))))+
      scale_x_discrete(labels=c("ambient" = "Ambient", "hot" = "Hot")) +
    ggtitle("Kruskal-Wallis, p = 0.020") +
      annotate("text", x = 1.45, y = 85, label = "*", size = 12) +
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=16, color="black"),
      axis.text=element_text(size=14, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=14)); total_fert_success_plot


ggsave(plot=total_fert_success_plot, "figures/reproductive_ecology/total_fertilization_success.png",height=8, width=8)
ggsave(plot=total_fert_success_plot, "figures/reproductive_ecology/total_fertilization_success.pdf",height=8, width=8)
```


Plot data as a mean and standard error plot.     
```{r}

total_fert_plot <- calculate_fert_success %>%
  group_by(treatment)%>%
  summarise(mean=mean(tot.fert.success, na.rm=TRUE), 
            n=length(tot.fert.success),
            se=sd(tot.fert.success, na.rm=TRUE)/sqrt(length(tot.fert.success)))%>%

   ggplot(., aes(x = treatment, y = mean, colour = treatment, fill = treatment)) +
     geom_bar(position="dodge", stat="identity") +
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se, group = treatment), colour = "black", width=.2,position=position_dodge(.9))+
    scale_colour_manual(name="Treatment", values=c("blue", "red")) +
    scale_fill_manual(name="Treatment", values=c("blue", "red")) +
    ylab(expression(bold(paste("Fertilization Success Rate (%)"))))+
    xlab(expression(bold(paste("Treatment"))))+
    scale_y_continuous(expand = c(0, 0), limits = c(0, 80)) +
    scale_x_discrete(labels = c("ambient" = "Ambient", "hot" = "Hot")) +
    theme_classic() + 
    #annotate("text", x = 1.4, y = 95, label = "*", size = 12) +
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=34),
      axis.text=element_text(size=30, color="black"), 
      legend.text=element_text(size=20),
      ); total_fert_plot

ggsave(plot=total_fert_plot, "figures/reproductive_ecology/total_fertilization_success_mean.png",height=20, width=10)
ggsave(plot=total_fert_plot, "figures/reproductive_ecology/total_fertilization_success_mean.pdf",height=20, width=10)
```

## Statistical analysis  

Run ANOVA test on larval size. 
```{r}
tot_fert_model<-aov(tot.fert.success~treatment, data=calculate_fert_success)
summary(tot_fert_model)
qqPlot(residuals(tot_fert_model))
leveneTest(residuals(tot_fert_model)~treatment, data=calculate_fert_success)
```

There appears to be a violation in variance assumptions but not normality. Conduct non-parametric test (Kruskal Wallis).  

```{r}
kruskal.test(tot.fert.success~treatment, data=calculate_fert_success)
``` 

There is a significant difference in fert success between groups with higher from hot treatment.

Kruskal-Wallis rank sum test

data:  tot.fert.success by treatment
Kruskal-Wallis chi-squared = 4.3226, df = 1, p-value = 0.03761



# Calculate the means of proportions across clevage stages

```{r}
#Select only the columns we need for each proportion in each stage 
cleav <- calculate_fert_success %>% 
  dplyr::select(tube.ID, treatment, replicate, egg_prop, two.cells_prop, four.cells_prop, more.cells_prop)%>%
  drop_na() #remove na's that could not be measured

#change from wide to long format for easier plots below
cleav_long <- gather(cleav, stage, tot.fert, egg_prop:more.cells_prop, factor_key=TRUE)

```


Present means and standard error of each group and save summary table.  

```{r}
cleav_long%>%
  group_by(stage, treatment)%>%
  summarise(n=length(tot.fert),
            Mean=format(round(mean(tot.fert), 3), 3), 
            SE=format(round(sd(tot.fert)/sqrt(length(tot.fert)),3),3))%>%
   dplyr::rename(Stage=stage, Treatment=treatment)%>%
  kbl(caption="Descriptive statistics of total fertilization success rate across developmental stage")%>%
  kable_classic(full_width=FALSE, html_font="Arial")%>%
  row_spec(0, bold = TRUE) 

#need to output to csv 
cleav_long%>%
  group_by(stage, treatment)%>%
  summarise(n=length(tot.fert),
            Mean=format(round(mean(tot.fert), 3), 3), 
            SE=format(round(sd(tot.fert)/sqrt(length(tot.fert)),3),3))%>%
  dplyr::rename(Stage=stage, Treatment=treatment)%>%
  write_csv(., "output/toal_fert_table.csv")
```

Plot data as a box plot.     
```{r}

#use PNW color palette, build palette
pal=pnw_palette("Moth", type = "discrete")

cleav_plot <- cleav_long %>%
    ggplot(., aes(x = treatment, y = tot.fert, color = stage)) +
    geom_boxplot(aes(colour=stage), lwd=1) +
    geom_point(aes(fill=stage), colour="black", pch = 21, size=4, position = position_jitterdodge(0.5), alpha=0.6) + 
    scale_colour_manual(name="Stage", values=pal, labels=c("One Cell", "Two Cells", "Four Cells", "> Four Cells" )) +
    scale_fill_manual(name="Stage", values=pal, labels=c("One Cell","Two Cells", "Four Cells", "> Four Cells" )) +
    ylab(expression(bold(paste("Fertilization Success Rate (%)"))))+
    xlab(expression(bold(paste("Treatment"))))+
    ggtitle("") +
    theme_classic() + 
    theme(
      legend.position="right",
      axis.title=element_text(face="bold", size=14),
      axis.text=element_text(size=12, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=12), 
      axis.text.x=element_text(angle=45, hjust=1)
      ); cleav_plot


ggsave(plot=cleav_plot, "figures/reproductive_ecology/cleav_fertilization_success.png",height=8, width=8)
ggsave(plot=cleav_plot, "figures/reproductive_ecology/cleav_fertilization_success.pdf",height=8, width=8)
```

Plot data as a mean and standard error plot.     
```{r}
cleav_plot_mean <- cleav_long %>%
  group_by(treatment, stage)%>%
  summarise(mean=mean(tot.fert, na.rm=TRUE), 
            n=length(tot.fert),
            se=sd(tot.fert, na.rm=TRUE)/sqrt(length(tot.fert)))%>%

    ggplot(., aes(x = stage, y = mean, fill = treatment)) +
     geom_bar(position="dodge", stat="identity") +
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.2,position=position_dodge(.9))+
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Fertilization Success Rate (%)"))))+
    xlab(expression(bold(paste("Stage"))))+
  scale_y_continuous(expand = c(0, 0), limits = c(0, 30)) +
  scale_x_discrete(labels=c("egg_prop" = "One Cell", "two.cells_prop" = "Two Cells", "four.cells_prop" = "Four Cells", "more.cells_prop" = "> Four Cells")) +
    theme_classic() + 
    #annotate("text", x = 1.4, y = 95, label = "*", size = 12) +
    theme(legend.position = c(.97, .99),
    legend.justification = c("right", "top"),
      legend.title=element_text(face="bold", size=30),
      legend.text=element_text(size=28),
      axis.title=element_text(face="bold", size=30),
      axis.text=element_text(size=30, color="black"), 
      strip.text.x=element_text(face="italic", size=20)
      ); cleav_plot_mean

ggsave(plot=cleav_plot_mean, "figures/reproductive_ecology/cleav_fert_success_mean.png",height=6, width=6)
ggsave(plot=cleav_plot_mean, "figures/reproductive_ecology/cleav_fert_success_mean.pdf",height=6, width=6)
```

Plot data as a mean and standard error plot, stacked bar plot.     
```{r}
#use PNW color palette, build palette
pal=pnw_palette("Moth", type = "discrete")

stacked_plot <- cleav_long %>%
    ggplot(., aes(x = treatment, y = tot.fert, fill = stage)) +
  geom_bar(stat="identity", position="fill") +
    scale_colour_manual(name="Cleavage", values=pal, labels=c("egg_prop" = "One Cell", "two.cells_prop" = "Two Cells", "four.cells_prop" = "Four Cells", "more.cells_prop" = "> Four Cells")) +
    scale_fill_manual(name="Cleavage", values=pal, labels=c("egg_prop" = "One Cell", "two.cells_prop" = "Two Cells", "four.cells_prop" = "Four Cells", "more.cells_prop" = "> Four Cells")) +
    ylab(expression(bold(paste("Proportion of embryos at each cleavage stage"))))+
    xlab(expression(bold(paste("Treatment"))))+
    scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(labels = c("ambient" = "Ambient", "hot" = "Hot")) +
    theme_classic() + 
  ggtitle("Cleavage") +
    theme(plot.title = element_text(face = "bold", size = 28, hjust = 0.5),
      legend.position="top",
      axis.title=element_text(face="bold", size=28),
      axis.text=element_text(size=28, color="black"), 
      legend.title=element_blank(),
      legend.text=element_text(size=28), 
      #axis.text.x=element_text(angle=45, hjust=1)
      ); stacked_plot

ggsave(plot=stacked_plot, "figures/reproductive_ecology/cleav_stacked_plot.png",height=20, width=10)
ggsave(plot=stacked_plot, "figures/reproductive_ecology/cleav_stacked_plot.pdf",height=20, width=10)
```


Join all plots for each normalization together.  

```{r}
fert_final <- plot_grid(total_fert_success_plot, stacked_plot, ncol=2, nrow=1, labels = c('A', 'B'),rel_heights= c(1,1,1,1), rel_widths = c(1,1,1,1.2), label_y=1, align="h", label_size = 60)

ggsave(filename="figures/reproductive_ecology/fert_clev_Figure.pdf", plot=fert_final, dpi=300, width=20, height=15, units="in")
ggsave(filename="figures/reproductive_ecology/fert_clev_Figure.png", plot=fert_final, dpi=300, width=20, height=15, units="in")
```

# **3. Eggs per bundle**  

## Import and manipulate data

Load in data.
```{r}
# Egg size data
bundle <- read_csv("data/reproductive_ecology/eggs_bundle.csv")

# select columns
bundle <- bundle %>%
  dplyr::select(colony, treatment, cohort, well, eggs.bundle)%>%
  drop_na() #remove na's that could not be measured

bundle$well<-as.factor(bundle$well)
bundle$colony<-as.factor(bundle$colony)
bundle$treatment<-as.factor(bundle$treatment)
bundle$cohort<-as.factor(bundle$cohort)

```

Remove na's from data frame.  
```{r}
bundle <- bundle %>%
  dplyr::select(colony, treatment, cohort, well, eggs.bundle)%>%
  drop_na() #remove na's that could not be measured
```


## Plot eggs per bundle   

Plot data with mean and standard error for each treatment 
```{r}
bundle %>%
  ggplot(aes(x = treatment, y = eggs.bundle, color = treatment)) +
  labs(x = "",y = "Eggs per Bundle") +
  geom_jitter(width = 0.1) +                                            # Plot all points
  stat_summary(fun.data = mean_cl_normal, fun.args = list(mult = 1),    # Plot standard error
               geom = "errorbar", color = "black", width = 0.5) +
  stat_summary(fun = mean, geom = "point", color = "black") + # Plot mean
  theme_classic()
```


Plot data as a box plot.     

```{r}
bundle_plot <- bundle %>%
    ggplot(., aes(x = treatment, y = eggs.bundle, colour=treatment)) +
    geom_boxplot(aes(colour=treatment), lwd=1) +
    geom_point(aes(fill=treatment), colour="black", pch = 21, size=4, position = position_jitterdodge(0.5), alpha=0.6) + 
    xlab("") + 
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Eggs per Bundle"))))+
  xlab(expression(bold(paste("Treatment"))))+
    scale_x_discrete(labels=c("ambient" = "Ambient", "hot" = "Hot")) +
   scale_y_continuous(expand = c(0, 0), limits = c(0, 20)) +
        ggtitle("one-way anova, p = 0.045") +
      annotate("text", x = 1.45, y = 18, label = "*", size = 12) +
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=16, color="black"),
      axis.text=element_text(size=14, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=14));bundle_plot

ggsave(plot=bundle_plot, "figures/reproductive_ecology/bundle.png",height=8, width=8)
ggsave(plot=bundle_plot, "figures/reproductive_ecology/bundle.pdf",height=8, width=8)
```



Plot data as a mean and standard error plot.     
```{r}

bundle_plot_mean <- bundle %>%
  group_by(treatment)%>%
  summarise(mean=mean(eggs.bundle, na.rm=TRUE), 
            n=length(eggs.bundle),
            se=sd(eggs.bundle, na.rm=TRUE)/sqrt(length(eggs.bundle)))%>%
    ggplot(., aes(x = treatment, y = mean)) +
    geom_point(aes(fill=treatment), colour="black", pch = 21, size=8, alpha=1) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se, group=treatment), width=0, linetype="solid", position=position_dodge(0.4), size=0.8, color="black")+
    xlab("") + 
    scale_colour_manual(name="Treatment", values=c("blue", "red"))+
    scale_fill_manual(name="Treatment", values=c("blue", "red"))+
    ylab(expression(bold(paste("Eggs per Bundle"))))+
   xlab(expression(bold(paste("Treatment"))))+
      ggtitle("one-way anova, p = 0.045") +
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5, vjust = -0.1),
      legend.position="none",
      axis.title=element_text(face="bold", size=16, color="black"),
      axis.text=element_text(size=14, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=14), 
      axis.text.x=element_text(angle=0, hjust=1, color="black")
      ); bundle_plot_mean

ggsave(plot=bundle_plot_mean, "figures/reproductive_ecology/bundle_size_mean.png",height=6, width=6)
ggsave(plot=bundle_plot_mean, "figures/reproductive_ecology/bundle_size_mean.pdf",height=6, width=6)
```

## Statistical analysis  

Run ANOVA test on eggs per bundle.
```{r}
bundle_model<-aov(eggs.bundle~treatment, data=bundle)
summary(bundle_model)
qqPlot(residuals(bundle_model))
leveneTest(residuals(bundle_model)~treatment, data=bundle)
```

Meets assumptions for normality and homogeneity of variance :) eggs per bundle are significantly higher in the heated parental history treatment.

              Df Sum Sq Mean Sq F value Pr(>F)  
treatment      1     31  30.507   4.012 0.0454 *
Residuals   1027   7809   7.604   


#combine eggs per bundle, egg size, and fertilization success

```{r}
#generate figure with larval survival

repro_plot <- bundle_plot | size_plot | total_fert_success_plot

repro_plot <- repro_plot + plot_annotation(tag_levels = 'A')

repro_plot

ggsave("output/reproductive_ecology/all_repro.png", plot = repro_plot, width = 18, height = 10)
ggsave("output/reproductive_ecology/all_repro.pdf", plot = temps, width = 20, height = 12)


```





